---
title: "Final Project"
author: "Elizabeth Chen & Benjamin Tompkins"
date: "5/6/2020"
output: html_document
---
## Introduction
Similar to common currencies, cryptocurrencies can be used to buy products and services. However, unlike other currencies, cryptocurrencies are digital and use cryptography to provide secure online transactions. Much of the interest in these unregulated currencies is to trade for profit even though cryptocurrencies can be in used regular purchase. Speculators at times drove prices of cryptocurrencies skyward.

## Preface and Background
Cryptocurrency is a form of payment that can be exchanged online for products and services. Many companies have issued their own cryptocurrencies, often called tokens, which can be used specifically to purchase the products and services provided by the company. Cryptocurrency can be considered similar to arcade tokens or casino chips, which need to be purchased or exchanged with real currency.

Cryptocurrencies use a technology called blockchain, which is a decentralized technology spread across many computers that manages and records transactions. One of the appealing characteristics of cryptocurrencies is its security.

According to CoinMarketCap.com, a market research website, more than 2,200 different cryptocurrencies are traded publicly. The total value of all cryptocurrencies as of June 2019 was estimated to be $246 billion. Since the cryptocurrency market is still fairly young, the current stock prices of cryptocurrencies can vary quite a bit. We hope to do some analysis on the following types of cryptocurrencies to see if we can predict some current market trends: Bitcoin, Litecoin and Ethereum.

For more information on how cryptocurrencies work: https://blockgeeks.com/guides/what-is-cryptocurrency/
For more information on cryptocurrency prices: https://cointelegraph.com/explained/how-cryptocurrency-prices-work-explained

Created in 2009 by Satoshi Nakamoto, Bitcoin was the first widely adopted cryptocurrency. Bitcoin not only uses peer-to-peer technology to operate with no central authority or banks, but also collectively manages transactions and bitcoin’s issuance. Bitcoin is open-source; its design is public, nobody owns or controls Bitcoin and everyone can take part. Nowadays, Bitcoin has become synonymous with cryptocurrency, however, this is not the only type of cryptocurrency available.
(https://medium.com/decryptionary/what-is-bitcoin-for-dummies-a-guide-for-beginners-8b3d9c0a8065)

Litecoin is one of the most prominent alternatives to Bitcoin and works upon the same fundamental principles. The transaction time of Litecoin is roughly two and a half minutes faster than that of Bitcoin. With four times as many Litecoins in circulation, it theoretically offers smaller divisions of coins to make smaller transaction values more feasible. Litecoin also uses a hashing algorithm, known as scrypt, which is supposed to keep Litecoin mining realistic for desktop users. Bitcoin uses the standard SHA256d algorithm which becomes more time as well as power intensive as time goes on. 
(https://www.forbes.com/sites/quora/2018/02/08/what-is-litecoin/#6b7bf4c333f7)
(https://www.wired.com/2013/08/litecoin/)

Ethereum is a cryptocurrency that took the technology behind bitcoin and expanded its capabilities. Ethereum is a decentralized network with its own interbet browser, coding language, and payment system. Etheruem ustilizes a peer-to-peer approach, where nodes download the Etheruem blockchain and enforce all the rules of the system, which allows the network to be honest and the nodes to recieve rewards. Ethereum seems to be a complex mode of cryptocurrency exchange. It will be interesting comparing it with other cryptocurrencies.
(https://cointelegraph.com/ethereum-for-beginners/what-is-ethereum)

## Evironment Setup

Libraries used to conduct our tutorial through the data science pipeline. 
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(caret)
library(broom)
library(pracma)
```

## Data Curation and Parsing

  We are using datasets from https://www.kaggle.com/sudalairajkumar/cryptocurrencypricehistory#bitcoin_price.csv. This dataset has multiple .csv files for each type of cryptocurrency. In this specific tutorial we will be looking at three of these .csv files corresponding to three different cryptocurrencies. The cryptocurrencies we will be looking at are: Bitcoin, Ethereum, and Litecoin. We chose these three in particular because these were the ones that we had heard of and were fairly popular. CoinMarketCap, a website that tracks the value and performance of various cryptocurrencies, has all three of the chosen cryptocurrencies of this particular tutorial all within the top 10.
  After downloading the dataset, we put each csv into its own table, then gave it an identifier of its type, then combined all three tables into one for easier data manipulation. All three datasets had the same column names making it easier to combine into something useful. We then used a lubridate function that made the Data column more useful for comparison and sorted by the date.

```{r load_data, warnings=FALSE}
bitcoin_tab <- read_csv("datasets/bitcoin_price.csv")
litecoin_tab <- read_csv("datasets/litecoin_price.csv")
ethereum_tab <- read_csv("datasets/ethereum_price.csv")
bitcoin_tab <- mutate(bitcoin_tab, coin_type="bitcoin")
litecoin_tab <- mutate(litecoin_tab, coin_type="litecoin")
ethereum_tab <- mutate(ethereum_tab, coin_type="ethereum")
crypto_tab <- bitcoin_tab %>%
  rbind(litecoin_tab) %>%
  rbind(ethereum_tab) %>%
  mutate(Date=mdy(Date)) %>% 
  arrange(desc(Date))

head(crypto_tab)
```

``` {r plots}
crypto_tab %>%
  ggplot(aes(x=Date, y=Open, color=coin_type)) +
    geom_line()  +
    labs(title = "Opening Price over Time",
         x = "Opening",
         y = "Date")

```


Plotting Closing value -Opening Value to see the best and worst days for each cryptocurrency. Based on the chart, The end of 2017 had a huge boom with the best daily differnces occuring in December 2017. In early 2018, however, all three cryptocurrencies experience their worst daily loss, signifying the event called the 2018 cryptocurrency crash, which occured from January 6, 2016 to February 6, 2018. In this period Bitcoin fell by about 65%, after an unprecedented boom in 2017. (en.wikipedia.org/wiki/2018_cryptocurrency_crash)
``` {r diff}
crypto_tab %>%
  mutate(diff=Close-Open) %>%
  group_by(coin_type) %>%
  arrange(desc(diff)) %>%
  ggplot(aes(x=Date, y=diff, color=coin_type)) +
  geom_line() +
  labs(title = "Difference in Closing price vs Opening Price over Time",
         x = "Difference",
         y = "Date")
```

Best days for bitcoin and litecoin occuring in December 2017.

``` {r best day}
crypto_tab %>%
  mutate(diff=Close-Open) %>%
  group_by(coin_type) %>%
  filter(diff==max(diff))
```

Bitcoin and Ethereum both experience huge losses in January of 2018, signifying the 2018 cryptocurrency crash.

```{r worst day}
crypto_tab %>%
  mutate(diff=Close-Open) %>%
  group_by(coin_type) %>%
  filter(diff==min(diff))
```

Trying to copy project 4 final_df doesnt work.
```{r prediction}
outcome_df <- crypto_tab %>%
   mutate(diff=Close-Open) %>%
   mutate(Direction = ifelse(diff>0, "up", "down"))
outcome_df

predictor_df <- outcome_df %>%
  filter(year(Date) < 2018)
predictor_df

standardized_df <- predictor_df %>%
  group_by(coin_type) %>%
  mutate(mean = mean(diff)) %>%
  mutate(sd = sd(diff)) %>%
  mutate(z = (diff - mean) / sd) %>%
  ungroup()
standardized_df

wide_df <- standardized_df %>%
  select(coin_type, Date, z) %>%
  tidyr::spread(Date, z)

matrix_1 <- wide_df %>%
  select(-coin_type) %>%
  as.matrix() %>%
  .[,-1]

matrix_2 <- wide_df %>%
  select(-coin_type) %>%
  as.matrix() %>%
  .[,-ncol(.)]

diff_df <- (matrix_1 - matrix_2) %>%
  magrittr::set_colnames(NULL) %>%
  as_data_frame() %>%
  mutate(coin_type = wide_df$coin_type)

final_df <- diff_df %>%
  inner_join(outcome_df %>% select(coin_type, Direction), by="coin_type") %>%
  mutate(Direction=factor(Direction, levels=c("down", "up"))) %>%
  select(-coin_type)
final_df
```

